Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-22 paper.tex
Good draft of mutations section

\subsection*{Simulating mutations}
\label{sec-mutations}
% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees.
The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format. Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
The output of the combined simulations of ancestry and mutations is
then some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences);
information about the mutations that have been simulated
is usually not retained or made available for subsequent analysis.

@@

The advantages of storing mutations rather than sequences have
recently become clear in the context of the vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format---consisting of a Newick tree
and associated mutations in a binary format---and
associated \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format, however) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustate the advantages of the general
approach of storing ancestry and mutations over sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.

@@

For version 1.0 we have greatly extended the sophistication of
\msprime's mutation generation engine, achieving near feature-parity
with \SeqGen.
We  support a large number of mutation models, including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
The new \texttt{sim\_mutations} function
(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
mutation model, but full support for the
original infinite sites model is also maintained.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Generating simulations in \msprime\ is extremely efficient.
Fig.~\ref{fig-mutations-perf} shows the time requried to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with $508,125$ edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge (see Fig.\ref{fig-ts-example}) basis, rather than
tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
we would have to consider $273,887,838$ branches
since there are $137,081$ trees and $1,998$ branches in each
tree ($\sim500$ times more work).

% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we consider a single tree, \msprime\ generates mutations
extremely efficiently.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim 260,000$ mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms
at this scale. It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

Things we can do: future work will need to do things like
Dawg~\citep{cartwright2005dna} and to implement models and
indels~\citep{fletcher2009indelible}.

>>>>

\subsection*{Simulating mutations}
\label{sec-mutations}
% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees.
The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format, and there are various
extensions to the basic nucleotide
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
The output of the combined simulations of ancestry and mutations is
then some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences);
information about the mutations that have been simulated
is usually not retained or made available for subsequent analysis.

@@

The advantages of storing mutations rather than sequences have
recently become clear in the context of the vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format (consisting of a Newick tree
and associated mutations in a binary format) and
the \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustrate the advantages of the general
approach of storing ancestry and mutations rather than sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.

@@

For version 1.0 we have greatly extended the sophistication of
\msprime's mutation generation engine, achieving near feature-parity
with \SeqGen.
We  support a large number of mutation models, including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
The new \texttt{sim\_mutations} function
(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
mutation model, but full support for the
original infinite sites model is also maintained.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Generating simulations in \msprime\ is extremely efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis (see Fig.~\ref{fig-ts-example}), rather than
tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
we would have to consider 273,887,838 branches
since there are 137,081 trees and 1,998 branches in each
tree ($\sim500$ times more work).

% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we consider a single tree, \msprime\ generates mutations
extremely efficiently.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms
at this scale. It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

There are many ways in which the mutation generation code
in \msprime\ could be extended by adapting existing methods.
For example, we hope to add support for
microsatellites~\citep{mailund2005coasim},
codon models~\citep{arenas2007recodon}
and indels~\citep{cartwright2005dna,fletcher2009indelible}.
